### 2018-06-23 paper/5-implementations.tex
Work on S5

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types:
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c is ib k v = Rebuilder c ib k v -> Build c is k v
type Rebuilder c i k v = k -> v -> Task c k v -> Task (MonadState i) k v
\end{minted}
So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
The \hs{Rebuilder} makes uses of the persistent build information \hs{ib}, while
the scheduler might augment that with further persistent information of its own,
yielding \hs{is}.

A \hs{Rebuilder} is a function which the \hs{Scheduler} uses to rebuild
a particular key, passing to it the key, the key's ``old'' value,
and a \hs{Task} that can compute the value of the key if necessary. It uses
the persistent build state \hs{i} (carried by the state monad) to decide
whether to re-build the value.  If doing so is unnecesary, it returns the
old value; otherwise it runs the supplied \hs{Task} to rebuild it.  In both
cases it updates the persistent build information \hs{i} to reflect what happened.
So a \hs{Rebuilder} wraps a \hs{Task c k v}, which unconditionally rebulds the key,
to make a \hs{Task (Monadstate i) k v}, which rebuilds the key only if necessary, and
does the necessary book-keeping. \simon{I'm puzzled by the absence of 'c' in the return type.}

These two abstractions are the key to modularity: \emph{we can combine any scheduler
  with any rebuilder, and obtain a correct build system}.
In this section we will write a scheduler for each column of Table~\ref{tab-build-systems},
and a rebuilder for each row; then combine them to obtain the build systems in the
body of the Table.

@@

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTime}: if the key is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the task's dependencies
(computed by \hs{dependencies}) are out of date.  If the key is
\hs{dirty}, we use `run task` to rebuild it, and update the state with
the new modification time of the key\footnote{The real \hs{make}
relies on the file system to track modification times, but we prefer
to make it explicit here.}; othewise we can just return the old vale.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, Map k [k]), Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (dirtyKeys, deps) <- get
    let dirty = key `Set.member` dirtyKeys ||
                case Map.lookup key deps of Nothing -> True
                                            Just ks -> any (`Set.member` dirtyKeys) ks
    if not dirty then return value else do
        put (Set.insert key dirtyKeys, deps)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad i (i, Chain k) k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    [@@]       = return [@@]
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State i (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-5mm}
\end{figure}

>>>>

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types:
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
The \hs{Rebuilder} makes uses of the persistent build information \hs{ir}, while
the scheduler might augment that with further persistent information of its own,
yielding \hs{i}.

A \hs{Rebuilder} is a function which the \hs{Scheduler} uses to rebuild a
particular key, passing to it the key, the key's current value, and a \hs{Task}
that can compute the value of the key if necessary. It uses the persistent build
information \hs{ir} (carried by the state monad) to decide whether to rebuild
the value. If doing so is unnecessary, it returns the current value; otherwise
it runs the supplied \hs{Task} to rebuild it. In both cases it can choose to
update the persistent build information \hs{ir} to reflect what happened. So a
\hs{Rebuilder} wraps a \hs{Task}~\hs{c}~\hs{k}~\hs{v}, which unconditionally
rebuilds the key, to make a \hs{Task}~\hs{(MonadState}~\hs{ir)}~\hs{k}~\hs{v},
which rebuilds the key only if necessary, and does the necessary book-keeping.
Note that the resulting \hs{Task} is always monadic; static dependency analysis
can be performed on the original \hs{Task}~\hs{Applicative} if need be.

These two abstractions are the key to modularity: \emph{we can combine any
scheduler with any rebuilder, and obtain a correct build system}.
In this section we will write a scheduler for each column of
Table~\ref{tab-build-systems}, and a rebuilder for each row; then combine them
to obtain the build systems in the table's body.

@@

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the key is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
(computed by \hs{dependencies}) are out of date. If the key is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the key\footnote{The real \Make
relies on the file system to track modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current value.
We assume that the external system, which invokes the build system, updates
\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, Map k [k]), Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (dirtyKeys, deps) <- get
    let dirty = key `Set.member` dirtyKeys ||
                case Map.lookup key deps of Nothing -> True
                                            Just ks -> any (`Set.member` dirtyKeys) ks
    if not dirty then return value else do
        put (Set.insert key dirtyKeys, deps)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store (ir, Chain k) k v) (Chain k)
    go _    [@@]       = return [@@]
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let newTask :: Task (MonadState ir) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State ir (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-5mm}
\end{figure}
